---
title: "Study_ERP012747_DADA2 "
author: "Bryshal Moore"
date: "2026-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Install binaries from Bioconductor version 3.16. See tutorial [DAADA2 Installation](https://benjjneb.github.io/dada2/dada-installation.html) and [dada2 tutorial](https://benjjneb.github.io/dada2/tutorial.html)

#Install BiocManager
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
```

#Install DADA2
```{r}
#BiocManager::install("dada2")
```
#Load the DADA2 library
```{r}

library(dada2); packageVersion("dada2")
library(ggplot2)
library(tidyverse)

```

```{r}
getwd()
```

```{r}
# Set the path to fastq files
all_path <- "/home/brymoore/wsg_gtdb_tax/pig_microbiome/study_ERP012747"

# Check filenames first
head(list.files(all_path))

studies <- list(
  study_A = list(
    pattern  = "^10296\\.A",
    region   = "V1-V3",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_B = list(
    pattern  = "^10296\\.B",
    region   = "V1-V2",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_C = list(
    pattern  = "^10296\\.C",
    region   = "V4",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_D = list(
    pattern  = "^10296\\.D",
    region   = "V3-V4",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_E = list(
    pattern  = "^10296\\.E",
    region   = "V4",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_F = list(
    pattern  = "^10296\\.F",
    region   = "V1-V3",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_G = list(
    pattern  = "^10296\\.G",
    region   = "V4",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_H = list(
    pattern  = "^10296\\.H",
    region   = "V4",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_I = list(
    pattern  = "^10296\\.I",
    region   = "V3-V5",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_J = list(
    pattern  = "^10296\\.J",
    region   = "V4",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_K = list(
    pattern  = "^10296\\.K",
    region   = "V4",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_L = list(
    pattern  = "^10296\\.L",
    region   = "V1-V3",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_M = list(
    pattern  = "^10296\\.M",
    region   = "V1-V3",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_N = list(
    pattern  = "^10296\\.N",
    region   = "V1-V3",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_O = list(
    pattern  = "^10296\\.O",
    region   = "V1-V3",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_P = list(
    pattern  = "^10296\\.P",
    region   = "V1-V2",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_Q = list(
    pattern  = "^10296\\.Q",
    region   = "V3",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_R = list(
    pattern  = "^10296\\.R",
    region   = "V1-V3",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_S = list(
    pattern  = "^10296\\.S",
    region   = "V3",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  ),
  study_T = list(
    pattern  = "^10296\\.T",
    region   = "V1-V3",
    truncLen = 240,
    trimLeft = 0,
    maxEE    = 2
  )
)
```

```{r}
# Run this first, then go back and fill in truncLen for each study
for (s_name in names(studies)) {
  s <- studies[[s_name]]
  
  fq_files <- sort(list.files(all_path, pattern = s$pattern, full.names = TRUE))
  
  if (length(fq_files) == 0) {
    cat("\n⚠️  No files found for", s_name, "at path:", all_path, "\n")
    cat("Files present:", paste(head(list.files(all_path)), collapse = ", "), "\n")
    next
  }
  
  cat("\nFound", length(fq_files), "files for", s_name, "\n")
  
  n_plot <- min(9, length(fq_files))
  sample_idx <- sample(length(fq_files), n_plot)
  
  p <- plotQualityProfile(fq_files[sample_idx])
  print(p + ggtitle(paste("Quality Profiles:", s_name, "-", s$region)))
}
```
In gray-scale is a heat map of the frequency of each quality score at each base position. The mean quality score at each position is shown by the green line, and the quartiles of the quality score distribution by the orange lines. The red line shows the scaled proportion of reads that extend to at least that position (this is more useful for other sequencing technologies, as Illumina reads are typically all the same length, hence the flat red line).


```{r filter-and-trim}
#Update truncLen in the studies list above before running this chunk
filtered_results <- list()

for (s_name in names(studies)) {
  s <- studies[[s_name]]
  
  fq_files <- sort(list.files(all_path, pattern = s$pattern, full.names = TRUE))
  
  if (length(fq_files) == 0) {
    cat("\n⚠️  No files found for", s_name, "\n")
    next
  }
  
  if (is.na(s$truncLen)) {
    cat("\n⚠️  truncLen is still NA for", s_name, "-- skipping. Set it after viewing quality plots.\n")
    next
  }
  
  sample_names <- sub("\\.fastq\\.gz$", "", basename(fq_files))
  
  filt_path <- file.path(all_path, "filtered", s_name)
  filtered  <- file.path(filt_path, paste0(sample_names, "_filt.fastq.gz"))
  names(filtered) <- sample_names
  
  out <- filterAndTrim(
    fq_files, filtered,
    truncLen    = s$truncLen,
    trimLeft    = s$trimLeft,
    maxEE       = s$maxEE,
    truncQ      = 2,
    rm.phix     = TRUE,
    compress    = TRUE,
    multithread = TRUE
  )
  
  cat("\n--- Filter summary:", s_name, "---\n")
  print(out)
  
  filtered_results[[s_name]] <- list(
    stats    = out,
    filtered = filtered[file.exists(filtered)]
  )
}
```
```{r learn-errors}
error_models <- list()

for (s_name in names(studies)) {
  
  if (is.null(filtered_results[[s_name]])) {
    cat("\n⚠️  No filtered results for", s_name, "-- skipping.\n")
    next
  }
  
  cat("\nLearning errors for:", s_name, "\n")
  
  filt_files <- filtered_results[[s_name]]$filtered
  err <- learnErrors(filt_files, multithread = TRUE)
  
  print(plotErrors(err, nominalQ = TRUE) + ggtitle(s_name))
  
  error_models[[s_name]] <- err
}
```
```{r}
dada_results  <- list()
derep_results <- list()

for (s_name in names(studies)) {
  
  if (is.null(filtered_results[[s_name]]) || is.null(error_models[[s_name]])) {
    cat("\n⚠️  Skipping", s_name, "-- missing filtered reads or error model.\n")
    next
  }
  
  cat("\nDenoising:", s_name, "\n")
  
  filt_files <- filtered_results[[s_name]]$filtered
  err        <- error_models[[s_name]]
  
  cat("Number of samples:", length(filt_files), "\n")
  
  derep <- derepFastq(filt_files)
  
  # If only one sample, derepFastq returns a single object — wrap it in a list
  if (class(derep) == "derep") {
    derep <- list(derep)
    names(derep) <- names(filt_files)
  } else {
    names(derep) <- names(filt_files)
  }
  
  dada_out <- dada(derep, err = err, multithread = TRUE)
  
  # Same fix for dada output
  if (class(dada_out) == "dada") {
    dada_out <- list(dada_out)
    names(dada_out) <- names(filt_files)
  }
  
  derep_results[[s_name]] <- derep
  dada_results[[s_name]]  <- dada_out
}
```

```{r}
# 1. Check filtered_results
length(filtered_results)
names(filtered_results)

# 2. Check error_models
length(error_models)
names(error_models)

# 3. Check what the safety check is actually seeing
for (s_name in names(studies)) {
  cat("\n", s_name, "\n")
  cat("filtered_results NULL?", is.null(filtered_results[[s_name]]), "\n")
  cat("error_models NULL?",     is.null(error_models[[s_name]]), "\n")
}
```

```{r denoise}
dada_results  <- list()
derep_results <- list()

for (s_name in names(studies)) {
  
  cat("\n--- Starting:", s_name, "---\n")
  
  filt_files <- filtered_results[[s_name]]$filtered
  err        <- error_models[[s_name]]
  
  cat("Number of filtered files:", length(filt_files), "\n")
  cat("Files exist:", all(file.exists(filt_files)), "\n")
  
  cat("Running derepFastq...\n")
  derep <- tryCatch({
    derepFastq(filt_files)
  }, error = function(e) {
    cat("derepFastq failed:", e$message, "\n")
    NULL
  })
  
  if (is.null(derep)) next
  
  if (class(derep) == "derep") {
    derep <- list(derep)
    names(derep) <- names(filt_files)
  } else {
    names(derep) <- names(filt_files)
  }
  
  cat("Derep successful. Samples:", length(derep), "\n")
  
  cat("Running dada...\n")
  dada_out <- tryCatch({
    dada(derep, err = err, multithread = TRUE)
  }, error = function(e) {
    cat("dada failed:", e$message, "\n")
    NULL
  })
  
  if (is.null(dada_out)) next
  
  if (class(dada_out) == "dada") {
    dada_out <- list(dada_out)
    names(dada_out) <- names(filt_files)
  }
  
  cat("Denoising successful for", s_name, "\n")
  
  derep_results[[s_name]] <- derep
  dada_results[[s_name]]  <- dada_out
}

cat("\nStudies successfully denoised:", length(dada_results), "\n")
cat("Names:", paste(names(dada_results), collapse = ", "), "\n")
```




```{r}
seq_tables <- list()

for (s_name in names(studies)) {
  
  if (is.null(dada_results[[s_name]])) {
    cat("\n⚠️  No dada results for", s_name, "-- skipping.\n")
    next
  }
  
  seqtab <- makeSequenceTable(dada_results[[s_name]])
  
  cat("\nAmplicon length distribution for", s_name, ":\n")
  print(table(nchar(getSequences(seqtab))))
  
  seq_tables[[s_name]] <- seqtab
}

```


Chimera Removal per Study
```{r}
seqtab_nochim <- list()

for (s_name in names(seq_tables)) {
  seqtab_nochim[[s_name]] <- removeBimeraDenovo(
    seq_tables[[s_name]],
    method      = "consensus",
    multithread = TRUE,
    verbose     = TRUE
  )
  
  cat("\n", s_name, "- Fraction retained after chimera removal:",
      round(sum(seqtab_nochim[[s_name]]) / sum(seq_tables[[s_name]]), 3), "\n")
}
```


Track Reads Per Study
```{r}
get_n <- function(x) sum(getUniques(x))

track_list <- lapply(names(studies), function(s_name) {
  
  if (is.null(filtered_results[[s_name]]) || is.null(dada_results[[s_name]])) {
    return(NULL)
  }
  
  raw_stats <- filtered_results[[s_name]]$stats
  dada_out  <- dada_results[[s_name]]
  
  # Denoised counts
  denoised <- if (class(dada_out) == "dada") {
    rep(get_n(dada_out), nrow(raw_stats))
  } else {
    sapply(dada_out, get_n)
  }
  
  # Chimera-removed counts
  nonchim <- if (!is.null(seqtab_nochim[[s_name]])) {
    rowSums(seqtab_nochim[[s_name]])
  } else {
    rep(NA, nrow(raw_stats))
  }
  
  data.frame(
    study    = s_name,
    sample   = rownames(raw_stats),
    input    = raw_stats[, 1],
    filtered = raw_stats[, 2],
    denoised = denoised,
    nonchim  = nonchim
  )
})

track_df <- do.call(rbind, Filter(Negate(is.null), track_list))
print(track_df)

# Also print percentage retained at each step
track_df$pct_filtered <- round(track_df$filtered / track_df$input * 100, 1)
track_df$pct_denoised <- round(track_df$denoised / track_df$input * 100, 1)
track_df$pct_nonchim  <- round(track_df$nonchim  / track_df$input * 100, 1)

print(track_df)
```

```{r}
names(seqtab_nochim)
length(seqtab_nochim)
```

```{r}
file.exists("/home/brymoore/wsg_gtdb_tax/curated_GTDB/GTDB_bac120_arc53_ssu_r226_genus.fna.gz")
```


```{r}
# ASVs per study
sapply(seqtab_nochim, function(x) length(getSequences(x)))

# Total if assigned separately (current approach)
sum(sapply(seqtab_nochim, function(x) length(getSequences(x))))

# Total if deduplicated (new approach)
length(unique(unlist(lapply(seqtab_nochim, getSequences))))
```

```{r}
# Get all unique ASVs across all studies
all_asvs <- unique(unlist(lapply(seqtab_nochim, getSequences)))
cat("Unique ASVs to assign:", length(all_asvs), "\n")

# Make a dummy seqtab with all unique ASVs
dummy_seqtab <- matrix(1L,
                        nrow = 1,
                        ncol = length(all_asvs),
                        dimnames = list("dummy", all_asvs))

# Assign taxonomy ONCE
cat("\nAssigning taxonomy on", length(all_asvs), "unique ASVs...\n")
taxa_r226_all <- assignTaxonomy(
  dummy_seqtab,
  "/home/brymoore/wsg_gtdb_tax/curated_GTDB/GTDB_bac120_arc53_ssu_r226_genus.fna.gz",
  tryRC       = TRUE,
  multithread = TRUE
)

# Add species ONCE
cat("\nAdding species...\n")
taxa_species_r226_all <- addSpecies(
  taxa_r226_all,
  "/home/brymoore/wsg_gtdb_tax/curated_GTDB/GTDB_bac120_arc53_ssu_r226_species_parsed_CLASS_SPP_CLD.fa.gz",
  tryRC   = TRUE,
  verbose = TRUE
)

# Map back to each study
taxa_r226         <- list()
taxa_species_r226 <- list()

for (s_name in names(seqtab_nochim)) {
  study_asvs <- getSequences(seqtab_nochim[[s_name]])
  
  taxa_r226[[s_name]]         <- taxa_r226_all[rownames(taxa_r226_all) %in% study_asvs, , drop = FALSE]
  taxa_species_r226[[s_name]] <- taxa_species_r226_all[rownames(taxa_species_r226_all) %in% study_asvs, , drop = FALSE]
  
  cat(s_name, ":", nrow(taxa_r226[[s_name]]), "ASVs mapped back\n")
}
```

```{r}
# Sanity check - mapped ASVs should match original counts
cat("\nSanity check:\n")
cat("Original ASV counts:\n")
print(sapply(seqtab_nochim, function(x) length(getSequences(x))))
cat("\nMapped back ASV counts:\n")
print(sapply(taxa_r226, nrow))
```


```{}
taxa_r226 <- list()

for (s_name in names(seqtab_nochim)) {
  
  cat("\nAssigning taxonomy for:", s_name, "\n")
  
  taxa_r226[[s_name]] <- assignTaxonomy(
    seqtab_nochim[[s_name]],
    "/home/brymoore/wsg_gtdb_tax/curated_GTDB/GTDB_bac120_arc53_ssu_r226_genus.fna.gz",
    tryRC      = TRUE,
    multithread = TRUE
  )
  
  cat("Done:", s_name, "-", nrow(taxa_r226[[s_name]]), "ASVs assigned\n")
}

# Check results for any study
head(taxa_r226[["study_A"]])
```

```{r}
file.exists("/home/brymoore/wsg_gtdb_tax/curated_GTDB/GTDB_bac120_arc53_ssu_r226_species_parsed_CLASS_SPP_CLD.fa.gz")
```

```{}
taxa_species_r226 <- list()

for (s_name in names(taxa_r226)) {
  
  cat("\nAdding species for:", s_name, "\n")
  
  taxa_species_r226[[s_name]] <- addSpecies(
    taxa_r226[[s_name]],
    "/home/brymoore/wsg_gtdb_tax/curated_GTDB/GTDB_bac120_arc53_ssu_r226_species_parsed_CLASS_SPP_CLD.fa.gz",
    tryRC      = TRUE,
    verbose    = TRUE
  )
  
  cat("Done:", s_name, "-", 
      sum(!is.na(taxa_species_r226[[s_name]][, "Species"])), 
      "ASVs assigned to species\n")
}
```

```{r}
# Check results
head(taxa_species_r226[["study_A"]])
```


```{}
# Create output directory
out_dir <- "/home/brymoore/wsg_gtdb_tax/pig_microbiome/study_ERP012747/results"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# 1. Write track reads summary
write.csv(track_df, 
          file.path(out_dir, "read_tracking.csv"), 
          row.names = FALSE)

# 2. Write seqtab_nochim per study (ASV counts per sample)
for (s_name in names(seqtab_nochim)) {
  write.csv(seqtab_nochim[[s_name]],
            file.path(out_dir, paste0(s_name, "_seqtab_nochim.csv")),
            row.names = TRUE)
}

# 3. Write taxonomy per study
for (s_name in names(taxa)) {
  write.csv(taxa[[s_name]],
            file.path(out_dir, paste0(s_name, "_taxonomy.csv")),
            row.names = TRUE)
}

# 4. Write taxonomy with species per study
for (s_name in names(taxa_species)) {
  write.csv(taxa_species[[s_name]],
            file.path(out_dir, paste0(s_name, "_taxonomy_species.csv")),
            row.names = TRUE)
}

# 5. Confirm all files were written
cat("\nFiles written to", out_dir, ":\n")
print(list.files(out_dir))
```


```{}
# Create output directory
out_dir <- "/home/brymoore/wsg_gtdb_tax/pig_microbiome/study_ERP012747/results"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# 1. Write track reads summary
write.csv(track_df, 
          file.path(out_dir, "read_tracking.csv"), 
          row.names = FALSE)

# 2. Write seqtab_nochim per study (ASV counts per sample)
for (s_name in names(seqtab_nochim)) {
  write.csv(seqtab_nochim[[s_name]],
            file.path(out_dir, paste0(s_name, "_seqtab_nochim.csv")),
            row.names = TRUE)
}

# 3. Write taxonomy per study
for (s_name in names(taxa)) {
  write.csv(taxa[[s_name]],
            file.path(out_dir, paste0(s_name, "_taxonomy.csv")),
            row.names = TRUE)
}

# 4. Write taxonomy with species per study
for (s_name in names(taxa_species)) {
  write.csv(taxa_species[[s_name]],
            file.path(out_dir, paste0(s_name, "_taxonomy_species.csv")),
            row.names = TRUE)
}

# 5. Confirm all files were written
cat("\nFiles written to", out_dir, ":\n")
print(list.files(out_dir))
```

---

```{r}
# Create output directory
out_dir <- "/home/brymoore/wsg_gtdb_tax/pig_microbiome/study_ERP012747/results"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# 1. Write track reads summary
write.csv(track_df,
          file.path(out_dir, "read_tracking.csv"),
          row.names = FALSE)

# 2. Write seqtab_nochim per study
for (s_name in names(seqtab_nochim)) {
  write.csv(seqtab_nochim[[s_name]],
            file.path(out_dir, paste0(s_name, "_seqtab_nochim.csv")),
            row.names = TRUE)
}

# 3. Write taxonomy per study
for (s_name in names(taxa_r226)) {
  write.csv(taxa_r226[[s_name]],
            file.path(out_dir, paste0(s_name, "_taxonomy_r226.csv")),
            row.names = TRUE)
}

# 4. Write taxonomy with species per study
for (s_name in names(taxa_species_r226)) {
  write.csv(taxa_species_r226[[s_name]],
            file.path(out_dir, paste0(s_name, "_taxonomy_species_r226.csv")),
            row.names = TRUE)
}

# 5. Write full deduplicated taxonomy tables
write.csv(taxa_r226_all,
          file.path(out_dir, "all_unique_asvs_taxonomy_r226.csv"),
          row.names = TRUE)

write.csv(taxa_species_r226_all,
          file.path(out_dir, "all_unique_asvs_taxonomy_species_r226.csv"),
          row.names = TRUE)

# Confirm everything written
cat("\nAll files written to:", out_dir, "\n")
print(list.files(out_dir))
```



### Written Out Files
```
results/
├── read_tracking.csv                    # all studies read counts
├── study_X_seqtab_nochim.csv            # ASV table study A
├── study_X_taxonomy.csv                 # taxonomy to genus study A
├── study_X_taxonomy_species.csv         # taxonomy with species study A

```
---


```{}
library(DECIPHER)

# Load the training set once outside the loop
load("~/wsg_gtdb_tax/GTDB_r226_classifier.RData")

ranks <- c("domain", "phylum", "class", "order", "family", "genus", "species")

taxa_decipher <- list()

for (s_name in names(seqtab_nochim)) {
  
  cat("\nRunning DECIPHER taxonomy for:", s_name, "\n")
  
  # Create DNAStringSet from ASVs
  dna <- DNAStringSet(getSequences(seqtab_nochim[[s_name]]))
  
  # Run IdTaxa
  ids <- IdTaxa(
    dna, 
    trainingSet, 
    strand     = "both", 
    processors = NULL,    # uses all available cores
    verbose    = TRUE
  )
  
  # Convert to matrix
  taxid <- t(sapply(ids, function(x) {
    m     <- match(ranks, x$rank)
    taxa  <- x$taxon[m]
    taxa[startsWith(taxa, "unclassified_")] <- NA
    taxa
  }))
  
  colnames(taxid) <- ranks
  rownames(taxid) <- getSequences(seqtab_nochim[[s_name]])
  
  taxa_decipher[[s_name]] <- taxid
  
  cat("Done:", s_name, "-", nrow(taxid), "ASVs assigned\n")
  cat("Species assigned:", sum(!is.na(taxid[, "species"])), "\n")
}
```


Have a chart that shows .

for the sktech did a breath filter to see 
make a table for singleton and excludin singletons


for next week bring he pollution metagenomes. 