---
title: "PRJNA266806_DADA"
author: "Bryshal Moore"
date: "2026-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



#Install binaries from Bioconductor version 3.16. See tutorial [DAADA2 Installation](https://benjjneb.github.io/dada2/dada-installation.html) and [dada2 tutorial](https://benjjneb.github.io/dada2/tutorial.html)

#Install BiocManager
```{r}
#if (!requireNamespace("BiocManager", quietly = TRUE))
 #   install.packages("BiocManager")
```

#Install DADA2
```{r}
#BiocManager::install("dada2")
```
#Load the DADA2 library
```{r}

library(dada2); packageVersion("dada2")
```
```{r}
getwd()

```


#Set the path to  fastq files
```{r}
path <- "/home/brymoore/wsg_gtdb_tax/pig_microbiome/PRJNA266806/fastq_files/trimmed" #Change me to the directory containing the fastq files

list.files(path)

# In side of the folder containing my fastq files
```

```{r}
# Single-end reads
fnFs <- sort(list.files(path, pattern = "_trimmed.fastq.gz", full.names = TRUE))

# Extract sample names by removing the .fastq.gz suffix
sample.names <- sapply(strsplit(basename(fnFs), "_trimmed.fastq.gz"), `[`, 1)

```



```{r}
sample.names
```

When looking at the fastq files, noticed that alot of the reads are really short ~50 bp long. Also seems to be of low quality from Mulitqc report.

```{r}
plotQualityProfile(fnFs[1:2])

```

```{r}
filt_path <- file.path(path, "filtered")
dir.create(filt_path)

filtFs <- file.path(filt_path, basename(fnFs))

out <- filterAndTrim(
  fnFs, filtFs,
  truncLen = 150,     # adjust from quality plot
  maxN = 2,
  maxEE = 2,
  truncQ = 2,
  rm.phix = TRUE,
  compress = TRUE,
  multithread = TRUE
)

```



```{r}
head(out)
```


```{r}

errF <- learnErrors(filtFs, multithread=TRUE)
plotErrors(errF, nominalQ=TRUE)

```


```{r}
derepFs <- derepFastq(filtFs)
names(derepFs) <- sample.names

```


```{r}
dadaFs <- dada(derepFs, err=errF, multithread=TRUE)

```

```{r}
seqtab <- makeSequenceTable(dadaFs)

dim(seqtab)
```
```{r}
# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
```

```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE)

```

```{r}
dim(seqtab.nochim)
```
```{r}
sum(seqtab.nochim)/sum(seqtab)
```


### Track reads through the pipeline
```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "nonchim")
rownames(track) <- sample.names
head(track)
```





### Add Tax DECIPHER

https://decipher.codes/Downloads.html
```{r}
library(DECIPHER)
```



```{r}
dna <- DNAStringSet(getSequences(seqtab.nochim)) # Create a DNAStringSet from the ASVs
load("/home/brymoore/wsg_gtdb_tax/curated_GTDB/GTDB_r226_classifier.RData") # CHANGE TO THE PATH OF YOUR TRAINING SET
ids <- IdTaxa(dna, trainingSet, strand="both", processors=NULL, verbose=TRUE) # use all processors
ranks <- c("domain", "phylum", "class", "order", "family", "genus", "species") # ranks of interest
#Convert the output object of class "Taxa" to a matrix analogous to the output from assignTaxonomy
taxid <- t(sapply(ids, function(x) {
        m <- match(ranks, x$rank)
       taxa <- x$taxon[m]
        taxa[startsWith(taxa, "unclassified_")] <- NA
        taxa
}))
colnames(taxid) <- ranks; rownames(taxid) <- getSequences(seqtab.nochim)
```


### Write out all of the taxonomy files

Save the file with the taxonomy from using DECIPHER
```{r}
getwd()
```


```{r}
#write.csv(taxid, "taxid.csv", row.names = T)
```

```{r}
#write.csv(seqtab.nochim, "sequence_table.csv", row.names = T)
```