---
title: "PRJNA257009_DADA"
author: "Bryshal Moore"
date: "2026-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Install binaries from Bioconductor version 3.16. See tutorial [DAADA2 Installation](https://benjjneb.github.io/dada2/dada-installation.html) and [dada2 tutorial](https://benjjneb.github.io/dada2/tutorial.html)

#### Install BiocManager
```{r}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
```

### Install DADA2
```{r}
#BiocManager::install("dada2")
```

### Load the DADA2 library
```{r}

library(dada2); packageVersion("dada2")
```

```{r}
getwd()
```

### Set the path to  fastq files
```{r}
path <- "/home/brymoore/wsg_gtdb_tax/pig_microbiome/PRJNA257009/fastq_files" #Change me to the directory containing the fastq files(unzipped)

list.files(path)

# In side of the folder containing my fastq files
```



```{r}
# Forward and reverse fastq.gz filenames have format: SAMPLENAME_1.fastq.gz and SAMPLENAME_2.fastq.gz
fnFs <- sort(list.files(path, pattern="_1.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_2.fastq.gz", full.names = TRUE))
# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq.gz
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
```

```{r}
sample.names
```


```{r}
data.frame(fnFs, fnRs)
```


```{r}
# Check lengths
length(fnFs)
length(fnRs)

# Check structure
str(fnFs)
str(fnRs)
```


### Inspect Quality Plots
```{r}
sample.names
```

```{r}
plotQualityProfile(fnFs[1:2])
```

```{r}
plotQualityProfile(fnRs[1:2])
```


### Mismatch Reads Check
```{r}
library(ShortRead)
```

```{r}
# Count reads in each file
countFastq <- function(f) length(readLines(gzcon(file(f, "rb")))) / 4

fwd_counts <- sapply(fnFs, countFastq)
rev_counts <- sapply(fnRs, countFastq)

```

```{r}

# Find mismatched pairs
mismatch_idx <- which(fwd_counts != rev_counts)
data.frame(
  forward = fnFs[mismatch_idx],
  fwd_n   = fwd_counts[mismatch_idx],
  rev_n   = rev_counts[mismatch_idx]
)
```


### Filter and Trim
Assign the filenames for the filtered fastq.gz files.
```{r}
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

```{r}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(0,0),
              maxN=2, maxEE=c(2,5), truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=FALSE) # On Windows set multithread=FALSE (only needed for filterAndTrim)
head(out)
```

```{r}

# See which samples are missing
fnFs_names <- gsub("_1.fastq.gz", "", basename(fnFs))
fnRs_names <- gsub("_2.fastq.gz", "", basename(fnRs))

```

```{r}
# Which R1 files have no matching R2
fnFs_names[!fnFs_names %in% fnRs_names]

# Which R2 files have no matching R1
fnRs_names[!fnRs_names %in% fnFs_names]
```


### Learn Error Rates

```{r}
errF <- learnErrors(filtFs, multithread=TRUE)
```


```{r}
errR <- learnErrors(filtRs, multithread=TRUE)
```


```{r}
plotErrors(errF, nominalQ=TRUE)
```


```{r}
plotErrors(errR, nominalQ=TRUE)
```


### Sample Inference
```{r}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
```

```{r}
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
```

```{r}
dadaFs[[1]] 
```

### Merge paired reads
```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# Inspect the merger data.frame from the first sample
head(mergers[[1]])
```


### Construct sequence table
```{r}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```

```{r}
# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
```

```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
```


```{r}
sum(seqtab.nochim)/sum(seqtab)
```


### Track reads through the pipeline
```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)
```



## Add Tax DECIPHER
```{r}
dna <- DNAStringSet(getSequences(seqtab.nochim)) # Create a DNAStringSet from the ASVs
load("/home/brymoore/wsg_gtdb_tax/curated_GTDB/GTDB_r226_classifier.RData") # CHANGE TO THE PATH OF YOUR TRAINING SET
ids <- IdTaxa(dna, trainingSet, strand="both", processors=NULL, verbose=TRUE) # use all processors
ranks <- c("domain", "phylum", "class", "order", "family", "genus", "species") # ranks of interest
#Convert the output object of class "Taxa" to a matrix analogous to the utput from assignTaxonomy
taxid <- t(sapply(ids, function(x) {
        m <- match(ranks, x$rank)
       taxa <- x$taxon[m]
        taxa[startsWith(taxa, "unclassified_")] <- NA
        taxa
}))
colnames(taxid) <- ranks; rownames(taxid) <- getSequences(seqtab.nochim)
```


# Write out all of the taxonomy files

Save the file with the taxonomy from using DECIPHER
```{}
write.csv(taxid, "taxid.csv", row.names = T)
```

```{}
write.csv(seqtab.nochim, "sequence_table.csv", row.names = T)
```