---
title: "PRJNA237795_DADA"
author: "Bryshal Moore"
date: "2026-04-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
getwd()
```


#Install binaries from Bioconductor version 3.16. See tutorial [DAADA2 Installation](https://benjjneb.github.io/dada2/dada-installation.html) and [dada2 tutorial](https://benjjneb.github.io/dada2/tutorial.html)

#### Install BiocManager
```{r}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
```

### Install DADA2
```{r}
#BiocManager::install("dada2")
```

### Load the DADA2 library
```{r}

library(dada2); packageVersion("dada2")
```

```{r}
library(DECIPHER); packageVersion("DECIPHER")
```


### Set the path to  fastq files
```{r}
path <- "/home/brymoore/wsg_gtdb_tax/pig_microbiome/PRJEB11368/fastq_files" #Change me to the directory containing the fastq files(unzipped)

list.files(path)

# In side of the folder containing my fastq files
```


```{r}
# Single-end reads
fnFs <- sort(list.files(path, pattern = ".fastq.gz", full.names = TRUE))

# Extract sample names by removing the .fastq.gz suffix
sample.names <- sapply(strsplit(basename(fnFs), ".fastq.gz"), `[`, 1)
```


```{r}
# Sanity checks
length(fnFs)
head(sample.names)
```

### Inspect Quality Plots

```{r}
plotQualityProfile(fnFs[1:2])
```


### Filter and Trim
Assign the filenames for the filtered fastq.gz files.
```{r}
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_filt.fastq.gz"))
names(filtFs) <- sample.names

```

```{r}
length(filtFs)
```


```{r}
out <- filterAndTrim(fnFs, filtFs, 
                     truncLen= 245,
                     maxEE= 2, 
                     truncQ=2, 
                     rm.phix=TRUE,
                     compress=TRUE, 
                     multithread= TRUE) # On Windows set multithread=FALSE (only needed for filterAndTrim)
head(out)
```

### Learn Error Rates

```{r}
errF <- learnErrors(filtFs, multithread=TRUE)
```

```{r}
plotErrors(errF, nominalQ=TRUE)
```


### Dereplicate
```{r}
derepFs <- derepFastq(filtFs)
```


# Sample inference - NO mergePairs step
```{r}
dadaFs <- dada(derepFs, err = errF, multithread = TRUE)
```

### Make Seq Table
```{r}
seqtab <- makeSequenceTable(dadaFs)
```


```{r}
dim(seqtab)
```

### Remove chimeras
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab, method = "consensus", multithread = TRUE)
```


```{r}
dim(seqtab.nochim)
```

```{r}
sum(seqtab.nochim)/sum(seqtab)
```

### Track reads through the pipeline
```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "nonchim")
rownames(track) <- sample.names
head(track)
```

https://decipher.codes/Downloads.html


### Add Tax DECIPHER
```{r}
dna <- DNAStringSet(getSequences(seqtab.nochim)) # Create a DNAStringSet from the ASVs
load("/home/brymoore/wsg_gtdb_tax/curated_GTDB/GTDB_r226_classifier.RData") # CHANGE TO THE PATH OF YOUR TRAINING SET
ids <- IdTaxa(dna, trainingSet, strand="both", processors=NULL, verbose=TRUE) # use all processors
ranks <- c("domain", "phylum", "class", "order", "family", "genus", "species") # ranks of interest
#Convert the output object of class "Taxa" to a matrix analogous to the utput from assignTaxonomy
taxid <- t(sapply(ids, function(x) {
        m <- match(ranks, x$rank)
       taxa <- x$taxon[m]
        taxa[startsWith(taxa, "unclassified_")] <- NA
        taxa
}))
colnames(taxid) <- ranks; rownames(taxid) <- getSequences(seqtab.nochim)
```

### Write out all of the taxonomy files

Save the file with the taxonomy from using DECIPHER
```{r}
#write.csv(taxid, "taxid.csv", row.names = T)
```

```{r}
#write.csv(seqtab.nochim, "sequence_table.csv", row.names = T)
```